---
title: "Activity 21: Hypothesis Testing"
author: "Your Name"
date: "December 7, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```


# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity21.Rmd` and `fev.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity21.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue.



# Part 1: loading packages and data

## Load packages

As always, we start by loading all the packages that we'll need. *Add `message = FALSE` on the first line of your code chunk to hide all the messages that get printed out when you load packages.*

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(broom) 
```


## Data context

Today we're going to return to our FEV data from the beginning of the semester. Recall that these data include information for a total of 654 children, ages 3--19. Each row in the dataset corresponds to a single child's visit to a doctor's clinic. Variables collected include:

- `age`: age (years)
- `fev`: forced expiratory volume (FEV) (liters per second); a measure of lung function
- `height`: height (inches)
- `sex`: binary sex (male/female)
- `smoke`: smoking status (smoker/nonsmoker) 


## Load data

Download `fev.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R and then look at the data to remind yourself what it looks like. *Add `message = FALSE` on the first line of your code chunk if you want to hide the message that gets printed out when you load data.*

```{r load-data}
smoking <- read_csv('fev.csv')
```


# Part 2: hypothesis testing 

Our scientific question of interest is: *does smoking decrease lung function in children?*

We'll perform a hypothesis test to answer this question. Remember the general hypothesis testing procedure from the *Stat 155 Notes*:

1. State hypotheses $H_0$ and $H_A$
2. Select a significance threshold $\alpha$ 
3. Calculate a test statistic $t_{obs} = \frac{estimate - null}{std \ error}$
4. Calculate the corresponding p-value $p = P(|T| \ge |t_{obs}| \mid H_0 \text{ is true})$
5. Make a decision  
    - If p-value $\le \alpha$, we reject $H_0$ in favor of $H_A$  
    - If p-value $> \alpha$, we fail to reject $H_0$ (remember that jurors' decisions are "guilty" and "not guilty", not "guilty" and "innocent")
    
Let's walk through these steps for the FEV data, assuming that we're answering this question using the simple linear regression model 

$$E[\text{FEV} \mid \text{smoke}] = \beta_0 + \beta_1 \text{smokesmoker}$$

## Steps 1 and 2: Hypotheses and Significance Threshold

We complete steps 1 and 2 before we look at our data.

**STOP AND DISCUSS:** stop here to discuss and answer Questions 1 & 2 with your group.



## Step 3: Test Statistic

Now, fit the linear regression model to our data. Save this model so you can use it later (e.g., `mod <- fev %>% with(lm(...))`), then look at the model summary.

```{r lm}
mod1 <- smoking %>% with(lm(fev ~ smoke))

summary(mod1)

```

**STOP AND DISCUSS:** stop here to discuss and answer Questions 3 - 5 with your group.



## Step 4: P-Value

Assuming that the null hypothesis is true, it turns out that our test statistic has a sampling distribution that is a bell curve centered around 0. We'll use this information to calculate our p-value: the probability of seeing a test statistic as or more extreme than the one we observed in our data, assuming that the null hypothesis is true. 

**STOP AND DISCUSS:** stop here to discuss and answer Questions 6 & 7 with your group.



## Step 5: Decision

Based on all of the above evidence, we can now make a decision about the hypotheses.

**STOP AND DISCUSS:** stop here to discuss and answer Questions 8 & 9 with your group.




# Part 3: another type of hypothesis test

It turns out there are multiple types of hypothesis tests that we could perform to answer our scientific question of interest: *does smoking decrease lung function in children?* 

In Part 2, we used linear regression and tested whether the smoker coefficient was equal to zero. Another commonly used test for this type of question is called a *t-test*. You may have seen this in other classes. Let's use a t-test to answer our scientific question...


## Step 1: Hypotheses

Our null and alternative hypotheses when we use a t-test are 

$$H_0: \mu_{smoker}  = \mu_{nonsmoker}$$  

$$H_A: \mu_{smoker} \neq \mu_{nonsmoker},$$ 

where $\mu_{group}$ represents the mean/average FEV for a given group. 

**STOP AND DISCUSS:** stop here to discuss and answer Question 10 with your group.

    
## Step 2: Significance Threshold

For comparison, use the same significance threshold for this hypothesis test as you did in Part 2. 
    

## Steps 3 and 4: Test Statistic and P-Value

We'll use the `t.test` function in R to get the test statistic and p-value for this t-test. Update the code below with the correct outcome variable (Y), the variable defining the groups we're comparing (X), and the name of your dataset (DATA). Find the test statistic (`t`) and p-value.

```{r ttest}
t.test(fev ~ smoke, data = smoking)
```


## Step 5: Decision

Based on all of the above evidence, we can now make a decision about the hypotheses.

**STOP AND DISCUSS:** stop here to discuss and answer Questions 11 & 12 with your group.
