---
title: 'Activity 3: Summarizing Data'
author: 'YOUR NAME'
date: 'November 2, 2020'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(eval = FALSE) # ignore unless you're an instructor and want to create a clean version for instructions
```


\

**STOP AND DISCUSS**: Before you get going in R/RStudio, stop here to discuss and answer Question 1 with your group.

\


# Getting started

1. Download `activity3.Rmd` and `introsurvey.csv` from Moodle and save them in a folder together on your computer that you can easily find again.  
   
2. Double check that the file you downloaded is called `activity3.Rmd`, not `activity3.Rmd.txt`.  
    
3. Open `activity3.Rmd` in RStudio: `File > Open File... >`.

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document called `activity3.html`.

**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue. 



# Part 1: loading R packages

As was the case in Activity 2, throughout this activity we're going to be using some *functions* that live inside R *packages*. We already *installed* some of those packages last class, but we will still need to *load* all of those packages.

1. Open the `Packages` panel (bottom right corner) to see the list of all packages that are already installed. Look to see if `stringr`,`readr`, `dplyr` and `ggplot2` are listed here. If they're on this list, that means they're already installed.

2. If all four packages are already installed, skip to Step 3. If any packages are missing from this list, type `install.packages('packagename')` in the *Console* (bottom left corner) and hit Enter/Return.

3. Load all packages by running the following code chunk (press the green "play" button in the top right corner of the code chunk). You should now see a check mark next to each of these packages in the `Packages` panel.

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
```

Remember that *installing* a package is like purchasing a book: we only need to **do it once** in order to get it on our bookshelf. *Loading* a package, on the other hand, is like taking a book off the bookshelf and opening it: we need to do this for **every R session** in which we want to use the functions in the package.



# Part 2: loading the Introductions Survey data

Now that we've installed and loaded all the packages that we need, we're ready to read our data into R. We'll be exploring your responses to the *Stat 155 Introductions Survey* that you completed on the first day of class.

Read the data into R by running the following code chunk:

```{r read-data}
results <- read_csv('introsurvey.csv')
```

If you get an error message here, remember these tips:

- read the error message to see if it gives you a hint about where the error occurred
- check your spelling, capitalization, punctuation, and order of operations
- make sure you've saved your .Rmd file
- check that the data are stored in the same folder as this .Rmd file (or that you've correctly specified the entire path to the data file)
- go back to Part 1 and make sure you loaded all necessary packages
- access the help page for the function you're trying to run (type `?functionname` into the Console) and/or copy and paste the error message into Google

If you've read in the data successfully, you should see a new data frame called `results` show up in the `Environment` panel (top right corner of your screen).



# Part 3: exploring the Introductions Survey data

Once your data are successfully loaded into R, look at the first six rows to remind yourself what the data look like: 

```{r look-at-data}
head(results)
```



# Part 4: summarizing data

Now that you've read in our data and performed some basic exploration, you're ready to start summarizing!

We use different techniques for summarizing data depending on the number and type of variables we're interested in studying. 


## Majors

The majors variable is *categorical*. We visualize the patterns in a categorical variable using a **bar plot**. 

To create a bar plot, start with the code below. You should see a blank canvas.

```{r barplot-step1}
results %>%
  ggplot(aes(x = `What is your declared or intended major? (undecided is a totally acceptable answer!)`))
```


Next, add a bar plot to the blank canvas. 

```{r barplot-step2}
results %>%
  ggplot(aes(x = `What is your declared or intended major? (undecided is a totally acceptable answer!)`)) + 
  geom_bar()
```


**STOP AND DISCUSS**: stop here to discuss and answer Question 2 with your group.


You might have noticed above that our bar plot is currently treating some similar responses as separate categories (e.g., "math" and "Mathematics"). This is a common challenge with open response questions. Run the code below to clean up the majors variable and create a bar plot of the newly cleaned variable. Do you think this is an improvement over the first version? 

```{r clean-majors-variable}
# create a new "clean" version of the majors variable (change to lower case and take first 4 characters)
results <- results %>%
  mutate(clean_major = trimws(str_sub(str_to_lower(`What is your declared or intended major? (undecided is a totally acceptable answer!)`),1,4)))

# plot the new clean major variable
# and rotate the x-axis labels so they don't overlap
results %>%
  ggplot(aes(x = clean_major)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))




```


## Birthdays

In Activity 2, you ran the code chunk below to count up how many people in our class will or will not be celebrating a birthday during Module 1.  

```{r summarize-birthdays}
results %>%
  count(`When is your birthday?`)
```

Run the code below to create a bar plot of this variable:

```{r visualize-birthdays}
results %>%
  ggplot(aes(x = `When is your birthday?`)) + 
  geom_bar()
```


**STOP AND DISCUSS**: stop here to discuss and answer Questions 3 and 4 with your group.


## Visualizing one quantitative variable

Work with your group to pick one *quantitative* variable from this dataset that you would like to summarize. We can visualize the patterns in quantitative variables using **histograms** or **boxplots**.

### Histograms

Histograms are constructed by (1) dividing up the observed range of the variable into 'bins' of equal width; and (2) counting up the number of cases that fall into each bin.

To create a histogram, start with the code below. Replace the question marks with the name of your variable. *If your variable has spaces or other special characters in its name, you'll need to put the name of the variable inside tick marks (e.g., `When is your birthday?`).*

```{r histogram-step1}
results %>%
  ggplot(aes(x = ??))
```

Next, add a histogram to the blank canvas. As before, you'll need to replace the question marks with the name of your variable. 

```{r histogram-step2}
results %>% 
  ggplot(aes(x = ??)) + 
  geom_histogram()


```


**STOP AND DISCUSS**: stop here to discuss and answer Question 5 with your group.


### Boxplots

Another way to visualize quantitative data is to create a boxplot. For more details, read [Section 2.6.3](https://bcheggeseth.github.io/Stat155Notes/one-quantitative-variable.html#boxplot) in the *Stat 155 Notes*.

Replace the question marks below with the name of your variable, then run the code to create a boxplot.

```{r boxplot}
results %>%
  ggplot(aes(y = ??)) + 
  geom_boxplot()
```


**STOP AND DISCUSS**: stop here to discuss and answer Question 6 with your group.


### Multiple Boxplots

Boxplots are very useful for comparing between groups. For more details, read [Section 2.7.2](https://bcheggeseth.github.io/Stat155Notes/one-quant-and-one-cat-variable.html#multiple-boxplots) in the *Stat 155 Notes*.

Replace the question marks `??` below with the name of your quantitative variable, then run the code to create multiple boxplots.

```{r multiple-boxplots}
results %>%
  filter(`What is your expected graduation year?` > 2020) %>% #removing instructor's responses :)
  ggplot(aes(x = factor(`What is your expected graduation year?`), 
             y = ??)) + 
  geom_boxplot()
```


**STOP AND DISCUSS**: stop here to discuss and answer Question 7 with your group.



# Part 5: Challenge!

Skip to the **Last step** section, below, if you don't have time to look at these challenge questions.


a. Adjust the code above to try create histograms of your chosen quantitative variable using different bin widths. What impact does it have on the visualization? What are the trade-offs in using small vs large bins? (Hint: try adding `binwidth = 0.5`, `binwidth = 50`, ... inside the `geom_histogram()` command.)

```{r histogram-change-bins, warning = FALSE, message = F}
## histogram with smaller bins

## histogram with larger bins

```

> REPLACE THIS TEXT WITH YOUR ANSWER



b. Another type of plot that's useful for summarizing quantitative variables is a **density plot**. Density plots are essentially smooth versions of histograms. Create a density plot for your chosen quantitative variable by using the code below. How does this compare to the histograms you made above? Which do you prefer? Why?


```{r density, warning = FALSE, message = F}
results %>%
  ggplot(aes(x = ??)) +  #change ?? to your variable name
  geom_density()
```

> REPLACE THIS TEXT WITH YOUR ANSWER



c. For the histogram below, add a title and more meaningful axis labels. Change the x- and y-axis labels on one of your plots. *Hint: do a Google search for something like "add axis labels ggplot".*

```{r add-title-and-axis-labels, warning = FALSE, message = F}
results %>%
  ggplot(aes(x = ??)) +  #change ?? to your variable name
  geom_histogram() # + add title & axis labels
```


d. Adjust the bar plot code below to color the histogram bars green. NOTE: color can be an effective tool... is it effective here?

```{r change-color}
# Make the bars green
results %>%
  ggplot(aes(x = ??)) + #change ?? to your variable name
  geom_histogram() #add color 
```   

> REPLACE THIS TEXT WITH YOUR ANSWER

e. Check out the Data Visualization [cheat sheet](https://rstudio.com/resources/cheatsheets/) or work through this RStudio primer on how to [Visualize Data](https://rstudio.cloud/learn/primers/3)



# Last step

Click `Knit` one last time and convert the `.html` file into a PDF by following these steps: 

- `Knit > Open in Browser > Print > Save as PDF` (Windows,Mac-Chrome)
- `Knit > Open in Browser > Export as PDF` (Mac-Safari). 

**Don't forget to check for error messages or formatting issues in your HTML/PDF file!** 
