---
title: "Activity 14: More Logistic Regression"
author: "Your Name"
date: "November 23, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```



# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity14.Rmd` and `voting18.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity14.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue.




# Loading Packages and Data

## Loading packages

As always, we start by loading all the packages that we'll need. 

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
```


## Data context

In this activity, we'll use the `voting18.csv` dataset, which comes from the VOTER (Views of the Electorate Research) Survey from 2018. The data collection was conducted in partnership with the survey firm YouGov. In total, 6,005 adults (ages 18 and up) with internet access took the survey online between April 5 and May 14, 2018.  The goal of the research group is to have more productive conversations where voters feel like they are truly heard. They hope the study group's research and analysis helps us understand each other and make our democracy more functional. A number of variables were collected on each adult, including:

- `turnout16_2016`: did the participant vote in the 2016 U.S. Election? (1 = yes, 2 = no)
- `wealth_2016`: participant's thoughts on the distribution of money and wealth in the U.S (1 = distribution is fair, 2 = should be more evenly distributed, 8 = don't know)
- `birthyr_2018`: participant's year of birth


Today, we'll use these data to investigate whether the birth year/age of the voters was associated with whether or not they voted in 2016. 


## Loading the data

Download `voting18.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R by running the following code chunk.

```{r load-data}
voters <- read_csv('voting18.csv')
```


Look at the first six rows and/or type `View(voters)` in the Console to remind yourself what the data look like.

```{r look-at-data}
head(voters)
```



# Part 1: summarizing the relationship between voting & birth year

In Activity 13, we explored whether economic beliefs are associated with voting. Today we'll investigate the relationship between *birth year* and voting turnout.

**STOP AND DISCUSS:** stop here to discuss and answer Question 1 with your group.


## Summarize

We'll create side-by-side boxplots plot to visually compare the birth year of participants that voted compared to those that did not. Write in code below to create a plot visualizing the relationship between `birthyr_2018` and `turnout16_2016`. Make sure that your code produces **multiple** boxplots.

```{r graphical-summary}
## your code here
```


When we are comparing one quantitative and one categorical variables, a useful numerical summary that we can provide to accompany a visualization like the one we created above is to calculate the center and spread of the quantitative variable within each category of the categorical variable. We can do this using the `group_by()` and `summarize()` functions in R. Update the code below to calculate the mean, median, standard deviation, and IQR of `birthyr_2018` within each group defined by `turnout16_2016`. You may need to remove missing values.

```{r numerical-summary}
## your code here
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 2 with your group.


# Part 2: modeling the relationship between voting & birth year

Before we can fit any regression models, we need to turn our categorical outcome variable `turnout16_2016` into an indicator variable. Right now, voter turnout is coded as 1 if the participant voted and 2 if they did not vote; we want to change this so that the turnout variable is 1 if they voted and 0 if they did not vote. Run the code below to create this updated `turnout16_2016` variable:

```{r create-indicator}
# recode the turnout16_2016 variable so it's 1/0 instead of 1/2 
voters <- voters %>%
  mutate(turnout16_2016 = ifelse(turnout16_2016 == 1, 1, 0))
```

## Linear regression model

Now that we have a binary (1/0) outcome, let's fit a linear regression model with `turnout16_2016` as the outcome and `birthyr_2018` as the predictor variable. When we fit a *linear* regression model to these data, we will be modeling the *probability* of voting as a function of participant birth year:

$$P[turnout16\_2016 = 1 \mid birthyr\_2018] = \beta_0 + \beta_1 birthyr\_2018$$

Enter code below to fit this linear regression model:

```{r linear-model}
## your code here
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 3 with your group.


## Logistic regression model

Next, we'll fit a logistic regression model to these data instead of a linear regression model. When we fit a *logistic* regression model to these data, we will be modeling the *log odds* of voting in 2016 as a function of participant birth year:

$$\log(Odds[turnout16\_2016 = 1 \mid birthyr\_2018]) = \beta_0 + \beta_1 birthyr\_2018$$

Enter code below to fit this logistic regression model.

```{r logistic-model}
## your code here
```


**STOP AND DISCUSS:** stop here to discuss and answer Question 4-6 with your group.


# Part 3: predicting voting based on birth year

We can use our models to predict whether or not a participant voted based on their birth year. Run the code below to use the `predict()` function to predict the probability of voting for participants with birth years of 1916, 1950, 1985, and 1998. Before you run this code, make sure that your linear regression model is saved as `mod1` and your logistic regression model is saved as `mod2`. 

```{r predict}
# prediction based on linear regression model
predict(mod1, newdata = data.frame(birthyr_2018 = c(1916, 1950, 1985, 1998)))

# prediction based on logistic regression model
predict(mod2, newdata = data.frame(birthyr_2018 = c(1916, 1950, 1985, 1998)), type = 'response')
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 7-9 with your group.

