---
title: "Activity 9: Multiple Linear Regression"
author: "Your Name"
date: "November 12, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```


# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.3) and RStudio (version 1.3.1093™): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity9.Rmd` and `housing.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   

    - I **strongly** recommend that you save this file in a new folder dedicated to activities for this class, *not in your Downloads*.
    - Double check that the file you downloaded does not have an extra `.txt` at the end of its name. 
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity9.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue. 



# Part 1: Load Packages and Data

## Loading packages

As always, we start by loading all the packages that we'll need. 

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
```

## Data context

This data in `housing.csv` describe the sale of individual residential properties in Ames, Iowa from 2006 to 2010. The data set contains 2930 observations and a large number of both categorical and quantitative variables involved in assessing home values. 

We will focus on three of these variables today:

- `SalePrice`: the sale price, in U.S. dollars
- `GrLivArea`: above grade (ground) living area, in square feet
- `OverallQual`: the overall material and finish quality (1-poor, 10-high)

## Loading the housing data

Download `housing.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R by running the following code chunk.

```{r load-housing-data}
housing <- read_csv('housing.csv')
```


Look at the first six rows and/or type `View(housing)` in the Console to familiarize yourself with what the data look like.

```{r look-at-housing-data}
head(housing)
```

# Part 2: SalePrice v. GrLivArea

To start, let's consider the relationship between the sale price (`SalePrice`) and the square footage of the house (`GrLivArea`).  

## Summarize

Let's create a scatterplot, and then calculate numerical summaries of the two variables individually, as well as calculate a measure of the strength of the linear relationship between them. 

```{r summarize-price-vs-livarea}
# create a plot
housing %>%
  ggplot(aes(x = GrLivArea, y = SalePrice)) +
  geom_point() +
  geom_smooth(se = FALSE, color = 'lightblue') +
  geom_smooth(method='lm', se = FALSE, color = 'steelblue') +
  theme_classic()

# calculate numerical summaries
housing %>%
  summarize(mean(SalePrice), median(SalePrice), sd(SalePrice), # summarize SalePrice
            mean(GrLivArea), median(GrLivArea), sd(GrLivArea), # summarize GrLivArea
            cor(SalePrice, GrLivArea)) # summarize relationship between the two
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 1 & 2 with your group. 


## Model 1

We will focus our models on homes that are smaller than 3000 square feet. Let's start by fitting a simple linear regression model with sale price and living area. We'll save this model as `mod1` so we can use it later.

```{r fit-mod1}
# fit model
mod1 <- housing %>% 
  filter(GrLivArea < 3000) %>% # temporarily leave out the large houses 
  with(lm(SalePrice ~ GrLivArea))

# look at coefficient estimates
mod1 %>%
  coef()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 3 & 4 with your group. 

Now, evaluate this model by creating a residual plot and calculating the standard deviation of the residuals, as well as R-squared. 

**STOP AND DISCUSS:** stop here to discuss and answer Question 5 & 6 with your group. 

```{r eval-mod1}
# insert code to create residual plot



# look at r-squared and sigma (sd of residuals)
glance(mod1)
```

## Model 2

Let's also consider incorporating overall quality of the home into the model. First, let's create a visualization that adds to the scatterplot we created earlier, now coloring the points according to different categories of home quality.

```{r viz-qual-cat}
housing %>%
  filter(GrLivArea < 3000) %>% # temporarily leave out the large houses
  mutate(QualCat = cut(OverallQual,4)) %>% # create a temporary quality categories variable (just for this plot because we didn't reassign housing, i.e., no <-)
  ggplot(aes(x = GrLivArea, y = SalePrice, color = QualCat)) + # color points by QualCat
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + # fit lines separately by QualCat
  scale_color_viridis_d('Overall Quality\n(1-poor, 10-high)') + # colorblind friendly color palette
  theme_classic()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 7 with your group. 

Now, let's fit (and evaluate) a model using both living area and a categorical version of overall home quality to predict home sale prices. 

```{r fit-mod2}
# fit model
mod2 <- housing %>% 
  filter(GrLivArea < 3000) %>% # leave out large houses
  mutate(QualCat = cut(OverallQual,4)) %>% # create QualCat variable
  with(lm(SalePrice ~ GrLivArea + QualCat))

# look at coefficient estimates
mod2 %>%
  coef()

# plot residuals versus fitted values
mod2 %>%
  augment() %>%
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point() +
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) +
  theme_classic()

# calculate model evaluation metrics
glance(mod2)

```

**STOP AND DISCUSS:** stop here to discuss and answer the rest of the questions with your group. 

