---
title: "Activity 20: Is the relationship real?"
author: "Your Name"
date: "December 3, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```


# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity20.Rmd` and `resume.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity20.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue.




# Part 1: loading packages and data

## Load packages

As always, we start by loading all the packages that we'll need. Note that we added `message = FALSE` on the first line of your code chunk to hide all the messages that get printed out when you load packages.

```{r load-packages, message = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(broom) 
library(mosaic)
```



## Data context

In this activity, we'll look at data from an experiment conducted in 2001-2002 that investigated the influence of race and gender on job applications. The researchers created realistic-looking resumes and then randomly assigned a name to the resume that "would communicate the applicant's gender and race" (e.g., they they assumed the name Emily would generally be interpreted as a white woman, whereas the name  Jamal would generally be interpreted as a black man).  They then submitted these resumes to job postings in Boston and Chicago and waited to see if the  applicant got a call back from the job posting. You can find a full description of the variables in this dataset [here](https://www.openintro.org/data/index.php?data=resume). To start, we'll focus on the following:

- `received_callback`: indicator that the resume got a call back from the job posting
- `firstname`: the first name used on the resume
- `race`: inferred race associated with the first name on the resume
- `gender`: inferred binary gender associated with the first name on the resume



## Load data

Download `resume.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R and then look at the data to familiarize yourself with what it looks like. Add `message = FALSE` on the first line of your code chunk if you want to hide the message that gets printed out when you load data.

```{r load-data}
resume <- read_csv('resume.csv')
```





# Part 2: logistic regression and confidence intervals

We'll use logistic regression to investigate the effect of inferred race (`race`) and binary gender (`gender`) on the odds of getting a callback (`received_callback`).


Fit this multiple logistic regression model in R using the `glm` function. Don't forget to specify `family = binomial`. 

```{r multiple-logistic-model}
mod1 <- resume %>%
  with(glm(received_callback ~ race + gender, family = binomial))

mod1 %>%
  coef()

expCoef <- mod1%>%
  coef() %>%
  exp()

expCoef

```

**STOP AND DISCUSS:** stop here to discuss and answer Questions 1 & 2 with your group.



Next, calculate 95% confidence intervals for the **exponentiated** coefficients.

```{r confint}
mod1%>%
  confint()%>%
  exp()
  
```

**STOP AND DISCUSS:** stop here to discuss and answer Questions 3-5 with your group.




# Part 3: logistic regression and hypothesis testing

The model that you fit in Part 2 can be written as 

$$\log(\text{Odds}[\text{Received Callback} \mid \text{race}, \text{gender}]) = \beta_0 + \beta_1 \text{raceWhite} + \beta_2 \text{genderMale}$$ 

Suppose we wanted to know whether inferred binary gender has a real effect on resume call backs, after adjusting for race. We could answer this question by looking at the 95% confidence interval you created in Part 2, or by testing the hypotheses 

$$H_0: \beta_2 = 0 \text{ vs } H_A: \beta_2 \neq 0,$$

or, equivalently,

$$H_0: e^{\beta_2} = 1 \text{ vs } H_A: e^{\beta_2} \neq 1.$$

**STOP AND DISCUSS:** stop here to discuss and answer Question 6 with your group.



Use the `tidy` function to get a clean version of the estimates from your fit logistic regression model. Look at the `p.value` column of the coefficient table. This column reports *p-values* for each of the coefficients in the model. 

```{r model-summary}
## your code here
```

**STOP AND DISCUSS:** stop here to discuss and answer Questions 7-10 with your group.



