---
title: "Activity 23: Model Selection"
author: "Your Name"
date: "December 9, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```


# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity23.Rmd` and `diamonds.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity23.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue.



# Part 1: loading packages and data

## Load packages

As always, we start by loading all the packages that we'll need. 

*Add `message = FALSE` on the first line of your code chunk to hide all the messages that get printed out when you load packages.*

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(broom) 
```


## Data context

Today we're going to explore a new data set about diamonds, a type of precious stone which is a piece of mineral crystal which, in cut and polished form, is used to make jewelry or other adornments. Diamonds are graded using a system developed by the Gemological Institute of America (GIA) in the early 1950s. They refer to the 4 C's (cut, clarify, color, carat) to help consumers better understand differences between stones. 

There are 10 variables in this dataset for over 50,000 individual diamonds. For today, we're going to focus on four of these: `carat`, `cut`, `color`, and `price`. Here's a brief description of each:

* `carat`: weight of the diamond (carats) (1 carat = 200 milligrams)
* `cut`: quality of the diamond cut (Fair, Good, Very Good, Premium, Ideal)
* `color`: diamond color, from best (D) to worst (J)
* `price`: price of the diamond (US dollars)


## Load data

Download `diamonds.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R and then look at the data to remind yourself what it looks like. 

*Add `message = FALSE` on the first line of your code chunk if you want to hide the message that gets printed out when you load data.*

```{r load-data}
diamonds <- read_csv('diamonds.csv')
head(diamonds)
```


```{r}
diamonds <- diamonds %>%
  mutate(log_price = log(price), log_carat = log(carat))
```

# Part 2: exploratory data analysis

## Explore relationships visually

We expect that you have little knowledge about diamonds or why they might be more or less expensive. In you group, divide the tasks to explore what aspects (carat, cut, color) may be most predictive of the price of a diamond. Remember that you may want to consider 3 variables (2 X's) at a time when looking at relationships with price. 

```{r create-visuals}
diamonds %>%
  ggplot(aes(x = depth, y = price)) + # color points by QualCat
  geom_point() + 
  geom_smooth() + # fit lines separately by QualCat
  scale_color_viridis_d('Overall Quality\n(1-poor, 10-high)') + # colorblind friendly color palette
  theme_classic()

diamonds %>%
  ggplot(aes(x = carat, y = price, color = color)) + # color points by QualCat
  geom_point() + 
  geom_smooth() + # fit lines separately by QualCat
  theme_classic()

diamonds %>%
  ggplot(aes(x = color, y = price)) + # color points by QualCat
  geom_boxplot() + 
  geom_smooth() + # fit lines separately by QualCat
  theme_classic()

diamonds %>%
  ggplot(aes(x = cut, y = price)) + # color points by QualCat
  geom_boxplot() + 
  geom_smooth() + # fit lines separately by QualCat
  scale_color_viridis_d('Overall Quality\n(1-poor, 10-high)') + # colorblind friendly color palette
  theme_classic()
```

Copy and paste your best visualizations to your group's Google Doc. 

```{r}
diamonds %>%
  ggplot(aes(x = log_carat, y = log_price, color = color)) + # color points by QualCat
  geom_point() + 
  geom_smooth() + # fit lines separately by QualCat
  theme_classic()
```


**STOP AND DISCUSS:** stop here to discuss and answer Questions 1-3 with your group.

# Part 3: model fitting and selection

## Fit Proposed Models

Now, let's fit the proposed models. First fit the simple linear model. 

```{r fit-simple-mod}
simple_mod <- diamonds %>%
  filter(!is.na(log_carat), !is.na(color), !is.na(log_price))
  with(lm(log_price ~ log_carat))
```

Now fit the multiple linear model you proposed as a group. 

```{r fit-multi-mod}
multi_mod <- diamonds %>%
  filter(!is.na(log_carat), !is.na(color), !is.na(log_price))
  with(lm(log_price ~ log_carat + color))
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 4 with your group.


## Compare Models

To compare these two models, let's first start by comparing R-squared and the standard deviation of residuals (residual standard error).

```{r eval-simple-mod}
glance(simple_mod)
```

```{r eval-multi-mod}
glance(multi_mod)
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 5 with your group.


Next, use a hypothesis test for the second predictor that you added to the multiple linear regression model to see if there is evidence that there is a relationship with price after accounting for the other variable. Remember what the null hypothesis is in this context?


```{r test-slope-multi-mod}
tidy(multi_mod)
```


Now, use a nested hypothesis test to compare the simple and multiple linear regression model to see if there is evidence that there the larger model is better. Remember what the null hypothesis is in this context?

```{r test-nested-models}
anova(simple_mod, multi_mod)
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 6 with your group.


## Check Conditions of Models

Before we choose a model, let's check to see if the conditions of the linear model are met with these two models. 

```{r condition-check-simple-mod}
# Replace with code to evaluate conditions of the simple model
```

```{r condition-check-multi-mod}
# Replace with code to evaluate conditions of the multiple linear model
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 7 & 8 with your group.


```{r play-space}
# Replace with code to consider relationships between X variables to create DAG
```


## Revise to get Final Model

Through this process of evaluating, comparing, and checking conditions of your models, you may decide to change the form of the variables (transforming, combining categories, including interactions) or which variables to include. In fact, model selection is an iterative process that requires you to 'go back to drawing board' (rethink the original model) to consider how best you can explain the variation in the outcome, here the price of a diamond.  


```{r new-multi-mod}
# Replace with code to revise the multiple linear model, you may add more than 2 variables
```

```{r play-space-two}
# Replace with code to justify choice of final model
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 9 with your group.

