---
title: "Activity 7: Residuals, Transformations, and R-squared"
author: "YOUR NAME"
date: "November 10, 2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
# knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```


# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity7.Rmd` and `fev.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   

    - I **strongly** recommend that you save this file in a new folder dedicated to activities for this class, *not in your Downloads*.
    - Double check that the file you downloaded does not have an extra `.txt` at the end of its name. 
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity7.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue. 



# Part 1: Load Packages and Data

## Loading packages

As always, we start by loading all the packages that we'll need. 

We are going to use one new package today, so before you run the code chunk below, you should first type `install.packages('broom')` in the Console.

```{r load-packages}
# load packages
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
```


## Data context (refresher)

Today we will continue working with the FEV data.

Forced expiratory volume (FEV) is a measure of lung capacity. Higher values indicate higher lung capacity---the ability to blow out more air in one second---and therefore indicate better lung function. Data are available for a total of 654 children, ages 3-19. Each row in the dataset corresponds to a single child's visit to a doctor's clinic. All children participating in this study attended the same pediatric clinic. The exact date of the study is unknown, but we at least know that it took place before 2005. Variables collected include:

- `age` at time of measurement (years)  
- measured forced expiratory volume `fev` (liters per second)  
- `height` at time of measurement (inches)  
- binary `sex` (male/female)  
- indicator of smoking habits `smoke` (smoker/nonsmoker)  


## Loading the FEV data

Download `fev.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R by running the following code chunk.

```{r load-fev-data}
fev <- read_csv('fev.csv')
```

Look at the first six rows and/or type `View(fev)` in the Console to remind yourself what the data look like.

```{r look-at-fev-data}
head(fev)
```




# Part 2: Checking model conditions (FEV vs height)

In the last activity, you fit a linear model for the expected FEV based on height of a child,

$$E[ FEV | height ] = \beta_0 + \beta_1 height$$

In this activity, we'll check model conditions (Section 3.8 in the Notes) and evaluate whether or not this is a "good" model (Section 3.7 in the Notes). 


## Look at scatterplot

One of the first things to do when checking model conditions is to create a scatterplot of the outcome variable (FEV) versus the predictor of interest (height) and check for **form** (is it *a straight line*,?) and **unusual features** (is there *equal spread* around that line? are there any *outliers*?) 

Run the code below to: 

- create a scatterplot of FEV versus height
- add the line of best fit (i.e., our linear regression line) in *steel blue*
- add another smooth "line" that is more flexible (i.e., not restricted to being a straight line) in *light blue*

```{r scatterplot}
fev %>%
  ggplot(aes(y = fev, x = height)) + 
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = 'steelblue') + # linear model
  geom_smooth(method = 'loess', se = FALSE, color = 'lightblue') +  # flexible smooth "line"
  theme_classic() # another 'clean' theme
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 1 & 2 with your group. 



## Calculating and plotting residuals

When checking model conditions it's also very useful to calculate and plot the model *residuals*, which tell us about how far our model is from the observed data points, in terms of the Y variable. 

First, we need to fit our model and save it. Below we call it `mod1` so it's easier to refer to later. Run this code to fit and save the model, and then print out the coefficients.

```{r model1}
# fit and save model
mod1 <- fev %>%
  with(lm(fev ~ height))

# print out coefficients
mod1 %>%
  coef()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 3 & 4 with your group. 


Now that you've thought about how to calculate the residual for a single data point, let's calculate the residuals for all observed data points in our dataset. We'll use the `augment` function to do this; it calculates the residuals, the predicted values, and a few other things that we are going to ignore for now, and it adds them as columns next to the variables used for the model. Run the code below, then double check for just the row or two that the residual (in the `.resid` column) is equal to the observed FEV value (in the `fev` column) minus the predicted/fitted FEV value (in the `.fitted` column). 

```{r augment-mod1}
augment(mod1) #broom package
```

Next, let's make two residual plots. Run the code below to plot the residuals against the fitted values, and then plot the residuals against the height of each child.

```{r resid-plots-mod1}
# residuals vs fitted values
augment(mod1) %>%
  ggplot(aes(y = .resid, x = .fitted)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + #se = FALSE removes gray interval
  geom_hline(yintercept = 0) +  #adds horizontal line as y = 0
  labs(x = 'Fitted Values', y = 'Residuals', title = 'Residuals vs Fitted Values') +
  theme_classic()

# residuals vs height
augment(mod1) %>%
  ggplot(aes(y = .resid, x = height)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) + 
  labs(x = 'Height', y = 'Residuals', title = 'Residuals vs Explanatory Variable') + 
  theme_classic()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 5 with your group. 



# Part 3: Transformations

## FEV versus cubed height

Let's try transforming our explanatory variable, height, and refitting our model. Run the code below to create a new `cubedheight` variable, plot FEV versus cubed height, and then fit a linear model for the expected FEV based on the cubed height of a child,

$$E[ FEV | height ] = \beta_0 + \beta_1 (height)^3 $$
```{r model2}
# create cubedheight variable
fev <- fev %>%
  mutate(cubedheight = height^3)

# plot FEV vs cubedheight
fev %>%
  ggplot(aes(y = fev, x = cubedheight)) + 
  geom_point() +
  theme_classic()

# fit and save model
mod2 <- fev %>%
  with(lm(fev ~ cubedheight))

# print out coefficients
mod2 %>%
  coef()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 6 with your group. 


Next, let's plot the residuals against the fitted values for this new model.

```{r resid-plots-mod2}
# residuals vs fitted values
augment(mod2) %>%
  ggplot(aes(y = .resid, x = .fitted)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) + 
  labs(x = 'Fitted Values', y = 'Residuals', title = 'Residuals vs Fitted Values (Model 2)') +
  theme_classic()


augment(mod2) %>%
  ggplot(aes(y = .resid, x = cubedheight)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) + 
  labs(x = 'Height Cubed', y = 'Residuals', title = 'Residuals vs Fitted Values (Model 2)') +
  theme_classic()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 7 with your group. 


## log FEV versus log height

Now, let's try a third model that transforms both our outcome, FEV, and our explanatory variable, height. Run the code below to create new `logFEV` and `logheight` variables, plot them, and then fit a linear model for the expected log FEV based on the log height of a child,

$$E[ \log(FEV) | height ] = \beta_0 + \beta_1 \log(height) $$
```{r model3}
# create logFEV and logheight variables
fev <- fev %>%
  mutate(logFEV = log(fev), logheight = log(height))

# plot logFEV vs logheight
fev %>%
  ggplot(aes(y = logFEV, x = logheight)) + 
  geom_point() +
  theme_classic()


# plot logFEV vs logheight
fev %>%
  ggplot(aes(y = logFEV, x = height)) + 
  geom_point() +
  theme_classic()

# fit and save model
mod3 <- fev %>%
  with(lm(logFEV ~ logheight))

# print out coefficients
mod3 %>%
  coef()
```

Next, let's plot the residuals against the fitted values for this third model.

```{r resid-plots-mod3}
# residuals vs fitted values
augment(mod3) %>%
  ggplot(aes(y = .resid, x = .fitted)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) + 
  labs(x = 'Fitted Values', y = 'Residuals', title = 'Residuals vs Fitted Values (Model 3)') +
  theme_classic()
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 8 with your group. 



# Part 4: Evaluating models

We've now fit three different linear regression models:

- Model 1: FEV versus height
- Model 2: FEV versus cubed height
- Model 3: log FEV versus log height

How can we decide which of these three models is best? 

To start, let's calculate summarize the residuals from these three models. Run the code below to calculate the mean and standard deviation of the residuals from each model.

```{r summarize-residuals}
# mean and sd of residuals for mod1
augment(mod1) %>%
  summarize(mean(.resid), sd(.resid))

# mean and sd of residuals for mod2
augment(mod2) %>%
  summarize(mean(.resid), sd(.resid))

# mean and sd of residuals for mod3
augment(mod3) %>%
  mutate(FEV = exp(logFEV),predFEV = exp(.fitted), residFEV = FEV - predFEV) %>%
  summarize(mean(.resid),mean(residFEV), sd(.resid),sd(residFEV))
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 9 with your group. 

Next, let's get the R squared ($R^2$) for all three models. Run the code below to use the `glance` function to get the R squared value, residual standard error  (`sigma`; ~ standard deviation of residuals), and other information for each model.

```{r compare-r2}
# get R^2 (and other info) for model 1
glance(mod1)

# get R^2 (and other info) for model 2
glance(mod2)

# get R^2 (and other info) for model 3
glance(mod3)
```
**STOP AND DISCUSS:** stop here to discuss and answer Question 10 & 11 with your group. 


# Wrapping Up

At the end of class (or after you've finished working through the questions above, whichever comes first), turn this .Rmd file into a nicely formatted .pdf as practice for your weekly assignments. To do this, follow these steps: `Knit` (to HTML) > `Open in Browser` > `Print` > `Save as PDF`. 
