---
title: 'Activity 8: Linear Regression with a Categorical Predictor'
author: "YOUR NAME"
date: "November 11, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
# knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```


# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity8.Rmd` and `fev.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   

    - I **strongly** recommend that you save this file in a new folder dedicated to activities for this class, *not in your Downloads*.
    - Double check that the file you downloaded does not have an extra `.txt` at the end of its name. 
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity8.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue. 




# Part 1: Load Packages and Data

## Loading packages

As always, we start by loading all the packages that we'll need. 

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
```


## Data context (refresher)

Today we will continue working with the FEV data, but this time we'll focus on building models with a *categorical* explanatory variable.

Forced expiratory volume (FEV) is a measure of lung capacity. Higher values indicate higher lung capacity---the ability to blow out more air in one second---and therefore indicate better lung function. Data are available for a total of 654 children, ages 3-19. Each row in the dataset corresponds to a single child's visit to a doctor's clinic. All children participating in this study attended the same pediatric clinic. The exact date of the study is unknown, but we at least know that it took place before 2005. Variables collected include:

- `age` at time of measurement (years)  
- measured forced expiratory volume `fev` (liters per second)  
- `height` at time of measurement (inches)  
- binary `sex` (male/female)  
- indicator of smoking habits `smoke` (smoker/nonsmoker)  


## Loading the FEV data

Download `fev.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R by running the following code chunk.

```{r load-fev-data}
fev <- read_csv('fev.csv')
```

Look at the first six rows and/or type `View(fev)` in the Console to remind yourself what the data look like.

```{r look-at-fev-data}
head(fev)
```


# Part 2: FEV vs smoking

To start, let's consider the relationship between lung function (`fev`) and smoking habits (`smoke`).  

**STOP AND DISCUSS:** stop here to discuss and answer Questions 1 & 2 with your group. 


## Summarize

Now that you've thought about what type of plot and numerical summaries you could make, let's create side-by-side boxplots and calculate the median, mean, and standard deviation of FEV in smoking and non-smoking children. 

```{r summarize-fev-vs-smoke}
# Create a boxplot with a clean theme_classic()
fev %>%
  ggplot(aes(x = smoke, y = fev)) +
  geom_boxplot() +
  labs(y = "FEV", title = "FEV for Smoking children vs. Non-Smoking Children") +
  theme_minimal()

# Calculate numerical summaries
fev %>%
  group_by(smoke) %>%
  summarize(mean(fev), sd(fev), median(fev))
```

**STOP AND DISCUSS:** stop here to discuss and answer Question 3 with your group. 


## Model

Now, let's fit a linear model for the expected FEV based on the smoking status of a child,

$$E[ FEV | smoke ] = \beta_0 + \beta_1 smokesmoker$$

**STOP AND DISCUSS:** stop here to discuss and answer Questions 4 & 5 with your group. 


```{r model-fev-vs-smoke}
fev %>%
  with(lm(fev ~ smoke))
```

**STOP AND DISCUSS:** stop here to discuss and answer Questions 6--8 with your group. 



# Part 3: FEV vs age categories

We can always *create* a categorical variable out of any of the quantitative variables in our dataset. Let's try creating a categorical variable out of age by separating children into age intervals.

```{r create-ageCat}
# Create new categorical variable with 4 age categories
fev <- fev %>%
  mutate(ageCat = cut(age, 4))

# Look at dataset with new variable added
head(fev)
```


## Summarize

Now that you've created this categorical age variable, let's create side-by-side boxplots and calculate the median, mean, and standard deviation of FEV in each age category. 

```{r summarize-fev-vs-ageCat}
# Create a boxplot


# Calculate numerical summaries

```

**STOP AND DISCUSS:** stop here to discuss and answer Question 9 with your group. 

## Model

Now, let's fit a linear model for the expected FEV based on the age category of a child,

$$E[ FEV | age ] = \beta_0 + \beta_1 ageCat(7,11] + \beta_2 ageCat(11,15] + \beta_3 ageCat(15,19]$$

```{r model-fev-vs-ageCat}
# Fit a linear regression model of FEV vs. age category

```

**STOP AND DISCUSS:** stop here to discuss and answer the remaining questions with your group. 


```{r summarize-smoke-vs-ageCat}
# Create a plot


# Calculate numerical summaries

```


# Wrapping Up

At the end of class (or after you've finished working through the questions above, whichever comes first), turn this .Rmd file into a nicely formatted .pdf as practice for your weekly assignments. To do this, follow these steps: `Knit` (to HTML) > `Open in Browser` > `Print` > `Save as PDF`. 

