---
title: "Activity 12: Model Selection"
author: "Your Name"
date: "November 18, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
# knitr::opts_chunk$set(eval = FALSE) # note for instructors: uncomment this code when creating a clean instructions version of the HTML; otherwise leave commented
```

# Getting Started

**If you have RStudio running and you've downloaded the Rmd and csv files from Moodle, skip to Part 1.**

Before you can work on this activity, you need access to RStudio. If you are struggling with any of the installation steps for R (version 4.0.2) and RStudio (version 1.3.1073): try using the [Macalester RStudio server](https://rstudio.macalester.edu) for this activity and then at the end of class today look at the *R and RStudio* section on Moodle and ask a group member and/or the instructor to troubleshoot the installation process.

Once you have access to RStudio, you're ready to start working on this activity:

1. Download `activity12.Rmd` and `housing.csv` from Moodle and save them someplace on your computer that you can find again (the two files need to be in the same folder together).   

    - I **strongly** recommend that you save this file in a new folder dedicated to activities for this class, *not in your Downloads*.
    - Double check that the file you downloaded does not have an extra `.txt` at the end of its name. 
  
2. Open RStudio.

3. Open your R Markdown file in RStudio: `File` > `Open File...` > find `activity12.Rmd` in your files. (If you're working on the Mac RStudio server, you need to upload the file first: go to the `Files` panel, then click `Upload`. Upload both the Rmd and the csv file.)

4. Update the author on the third line of the file.

5. Click the `Knit` button (look for the yarn and needle towards the top fo the screen). If successful, you should have a new window pop up with a nice looking HTML document. 


**Ask for help** if you encounter issues on any of the steps above. Once you've successfully made it through these steps, you can continue.




# Loading Packages and Data

## Loading packages

As always, we start by loading all the packages that we'll need. 

```{r load-packages}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(GGally)
library(forcats) #install.packages('forcats')
source('ggavplot.R') #install.packages('car'); install.packages('gridExtra')
```


## Data context

This data in `housing.csv` describe the sale of individual residential properties in Ames, Iowa from 2006 to 2010. The data set contains 2930 observations and a large number of both categorical and quantitative variables involved in assessing home values. 

We will focus on nine of these variables today:

- `SalePrice`: the sale price, in U.S. dollars
- `GrLivArea`: above grade (ground) living area, in square feet
- `OverallQual`: the overall material and finish quality (1-poor, 10-high)
- `OverallCond`: the overall condition of the home (1-poor, 10-high)
- `YearBuilt`: the year the home was originally built
- `FullBath`: the number of full baths (with bathtub)
- `HalfBath`: the number of half baths (toilet and sink)
- `BedroomAbvGr`: the number of bedrooms above ground
- `SaleCondition`: the condition of the sale (Normal, Abnormal, Partial, Family, etc.)


## Loading the housing data

Download `housing.csv` from Moodle (if you haven't already) and make sure it's saved in the same folder as this .Rmd file. Load the dataset into R by running the following code chunk.

```{r load-housing-data}
housing <- read_csv('housing.csv')
```


Look at the first six rows and/or type `View(housing)` in the Console to remind yourself what the data look like.

```{r look-at-housing-data}
head(housing)
```




# Part 1: Added Variable Plot & Redundancy

In Activity 11, we looked at the relationship between sale prices (`SalePrice`), above-ground living area (`GrLivArea`), and the overall quality (`OverallQual`) of homes in Ames, Iowa. 

Today we are going to look at more variables. 


## Fit large model

Run the code chunk below to create two new categorical variables: `QualCat` (categorical home quality) and `CondCat` (categorical home condition). These categorical variables will have 3 categories, and we'll also specify the reference category. 

For `QualCat`, the categories will be named "low-quality" (overall quality between 1 and 3), "mid-quality" (overall quality between 4 and 7), and "high-quality" (overall quality between 8 and 10), and we'll set low-quality homes as the reference category.

For `CondCat`, the categories will be named "poor-condition" (overall condition between 1 and 3), "good-condition" (overall condition between 4 and 7), and "great-condition" (overall condition between 8 and 10), and we'll set poor-condition homes as the reference category. 


```{r create-variables}
housing <- housing %>%
  filter(GrLivArea< 3000) %>%
  mutate(QualCat = relevel(factor(case_when(
    OverallQual %in% c(1, 2, 3) ~ "low-quality",
    OverallQual %in% c(4, 5, 6, 7) ~ "mid-quality",
    OverallQual %in% c(8, 9, 10) ~ "high-quality")),
    ref="low-quality")) %>%
  mutate(CondCat = relevel(factor(case_when(
    OverallCond %in% c(1, 2, 3) ~ "poor-condition",
    OverallCond %in% c(4, 5, 6, 7) ~ "good-condition",
    OverallCond %in% c(8, 9, 10) ~ "great-condition")),
    ref="poor-condition"))
```


Type `View(housing)` in the Console to look at the data set, which should now have a new categorical variable called `QualCat`. 



## Fit and check fullmod

Now, let's fit a linear regression model that models the average sale price as a function of home quality, home condition, year built, square footage, number of bedrooms, and number of full bathrooms. We'll save this model as `fullmod` so we can refer back to it later.

```{r fullmod}
# fit model and save as fullmod
fullmod <- housing %>%
  with(lm(SalePrice ~ QualCat + CondCat + GrLivArea + YearBuilt + BedroomAbvGr + FullBath))

# print out coefficients for fullmod
fullmod %>%
  coef()
```


Create an added variable plot for this model by running the code below:

```{r fullmod-avplot}
fullmod %>% 
  ggAVPLOTS()
```


And look for patterns in residuals:

```{r residuals}
augment(fullmod) %>%
  left_join(housing) %>%
  ggplot(aes(x = fct_reorder(Neighborhood,.resid), y = .resid)) +
  geom_boxplot() +
  geom_hline(yintercept = 0)+
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = 'Neighborhood',y = 'Residuals')
```


## Explore relationships between variables

In building a model it's also helpful to explore the relationships between the variables we're considering. What do we learn from these plots?

```{r var-explore}
housing %>%
  ggplot(aes(x = QualCat, fill = CondCat)) + 
  geom_bar(position='fill') +
  theme_classic() +
  scale_fill_viridis_d()


housing %>%
  select(GrLivArea, YearBuilt, BedroomAbvGr,FullBath) %>%
  ggpairs()


housing %>%
  ggplot(aes(x = QualCat, y = GrLivArea)) + 
  geom_boxplot() +
  theme_classic() +
  scale_fill_viridis_d() 

housing %>%
  ggplot(aes(x = CondCat, y = GrLivArea)) + 
  geom_boxplot() +
  theme_classic() +
  scale_fill_viridis_d() 

housing %>%
  ggplot(aes(x = QualCat, y = YearBuilt)) + 
  geom_boxplot() +
  theme_classic() +
  scale_fill_viridis_d() 


housing %>%
  ggplot(aes(x = CondCat, y = YearBuilt)) + 
  geom_boxplot() +
  theme_classic() +
  scale_fill_viridis_d() 

```




